### 2011-12-19 paper_draft.tex

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{openSNP - Crowdsourcing Genome Wide Association Studies}
}
% Insert Author names, affiliations and corresponding author email.
\\
Bastian Greshake$^{1,\ast}$, 
Philipp Bayer$^{2}$, 
Fabian Zimmer$^{3,}$
\\
\bf{1} Bastian Greshake Frankfurt am Main, Germany
\\
\bf{2} Philipp Bayer Gold Coast, Australia
\\
\bf{3} Fabian Zimmer M\"unster, Germany
\\
$\ast$ E-mail: info@opensnp.org
\end{flushleft}

@@

Genome Wide Association Studies (GWAS) are an easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which can be interesting because of their medical relevance. SNPs found through GWAS can be used to find candidate genes for a closer inspection or to predict disease risks. Genome Wide Association Studies make use of statistics to compare the alleles of patients to the alleles of healthy controls. By this the method does not allow to find causal differences but mere correlations. The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group (doi:10.1126/science.1109557). Since the beginning the number of participants in those studies is rising and over 1200 GWAS have been performed (doi:10.1186/1471-2350-10-6.) and over 5000 SNPs have been linked to different diseases and traits in those studies %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table).

@@

On the other hand those hand those people who are not planning to get genotyped are more likely to not share their data, because they agree to fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they agree that they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because agree that they fear negative consequences for their family (mean difference = 0.733, SE = 0.21, p = 0.002) or because they agree that they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).

@@

\subsection*{openSNP}
We've created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints. Users are encouraged to list as many phenotypes as possible through to a simple achievement-system which rewards users that upload their data and enter phenotypic information with small badges for their efforts. The possible answers in terms of variations for a single phenotype are not limited and every user can add completely new phenotypes if the corresponding questions about this are lacking.

@@

Between the start of openSNP on 09/29/2011 and 12/17/2011 214 people have signed up with openSNP, 79 of those have uploaded their genotyping results. In the same timeframe all users combined have entered 675 variations which are distributed over 47 different phenotypes. The phenotype which has been provided by the most users is the eye color which has been provided by 54 different users. There are two phenotypes which have only been provided by a single user: The score of the Writing-SAT and triglyceride-levels.

@@

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Survey on Sharing Genetic Information}
The survey was done with Google Docs and included questions on the age, chromosomal sex and ethnicity of the participants. Additionally it included questions on if they are already customer of a DTC company, are planning to become one or don't plan to become one. If they are already a customer they also got asked if they already share their genetic and phenotypic data. All participants got asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".

@@

\subsection*{Technical realization of the platform}
The main platform is realized using the web-framework Ruby on Rails. Postgres is used as the main database backend for Rails. The database stores genotyping results, phenotypic information of the users, literature results of Mendeley, the Public Library of Science and summaries on SNPs which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API.

SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number. This is a common format which is used by the NCBI dbSNP database and is also widely used and easily parsed from different literature-sources.

@@

The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{openSNP - Crowdsourcing Genome Wide Association Studies}
}
% Insert Author names, affiliations and corresponding author email.
\\
Bastian Greshake$^{1,\ast}$, 
Philipp Bayer$^{2}$, 
Fabian Zimmer$^{3}$,
Julia Reda$^{4}$
\\
\bf{1} Goethe University, Frankfurt am Main, Germany
\\
\bf{2} Bond University, Gold Coast, Australia
\\
\bf{3} Westf\"alische Wilhelms Universit\"at, M\"unster, Germany
\\
\bf{4} Johannes-Gutenberg University, Mainz, Germany
\\
$\ast$ E-mail: info@opensnp.org
\end{flushleft}

@@

Genome Wide Association Studies (GWAS) are a comparatively easy and cheap way to find Single Nucleotide Polymorphisms (SNPs) which can be interesting because of their medical relevance. SNPs found through GWAS can be used to find candidate genes for a closer inspection or to predict disease risks. Genome Wide Association Studies make use of statistics to compare the alleles of patients to the alleles of healthy controls. By this the method does not allow to find causal differences but mere correlations. The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group (doi:10.1126/science.1109557). Since the beginning the number of participants in those studies is rising and over 1200 GWAS have been performed (doi:10.1186/1471-2350-10-6.) and over 5000 SNPs have been linked to different diseases and traits in those studies %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table).

@@

On the other hand those people who are not planning to get genotyped are more likely to not share their data, because they agree to fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they agree that they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because agree that they fear negative consequences for their family (mean difference = 0.733, SE = 0.21, p = 0.002) or because they agree that they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).

@@

\subsection*{openSNP}
We created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints. Users are encouraged to list as many phenotypes as possible through to a simple achievement-system which rewards users that upload their data and enter phenotypic information with small badges that are shown on their profile page.

The possible answers in terms of variations for a single phenotype are not limited and every user can add completely new phenotypes if the corresponding questions about this are lacking. To reduce the amount of manual data curation openSNP tries to avoid the entry of the same phenotype or variation, but with a slightly different spelling by helping users at entering data by an autocompletion-feature which lists similar entries which are already in the openSNP-database.

@@

Between the start of openSNP on 09/27/2011 and 12/18/2011 214 people have signed up with openSNP, 79 of those have uploaded their genotyping files. Through this the openSNP database lists 69486471 SNPs which are distributed over 1938604 unique Rs-IDs. In the same timeframe all users combined have entered 675 variations which are distributed over 47 different phenotypes. See figure n for a distribution of data acquisition over time.

The mean number of users that have entered their variation for a single phenotype is 14.36 (SD 12.65), the median is 10. The distribution of how many users have entered their data per phenotype can be seen in figure n+1. The phenotype which has been provided by the most users is the eye color which has been provided by 54 different users. There are two phenotypes which have only been provided by a single user: The score of the Writing-SAT and triglyceride-levels.

A total number of 15229 documents relevant to the SNP-IDs which are listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. Of the primary literature 25 \% are released in Open Access-journals and can be freely free of charge by every user (Figure n+2). For usability reasons SNPs are ranked by the amount of information gathered by the external services.

The external services themselves are ranked by how easy users can access information out of these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of the Public Library of Science. Lowest values are given to the Mendeley-results, as those aren't necessary freely available for every user. SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley-entry.

@@

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Survey on Sharing Genetic Information}
The survey was done with Google Docs and was distributed to possible participants through the 23andMe-community, the DIYBiology mailinglist, blogs which focus on genetics and DTC genetic testing and social media like Twitter, Google Plus and Facebook. Because of this the results are biased in a way that customers of DTC genetic testing are overrepresented. As we wanted to find out how people who have purchased DTC genetic testing care about sharing their data with third parties this does not lead to further problems.

The survey included questions on the age, chromosomal sex and ethnicity of the participants. Additionally it included questions on if they are already customer of a DTC company, are planning to become one or don't plan to become one. If they are already a customer they also got asked if they already share their genetic and phenotypic data. All participants got asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".

@@

\subsection*{Technical realization of the platform}
The main platform is realized using the web-framework Ruby on Rails. Postgres is used as the main database backend for Rails. The database stores genotyping results, phenotypic information of the users, literature results of Mendeley, the Public Library of Science and summaries on SNPs which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API. All databases are queried using the unique identifier of each SNP as search term.

SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number. This is a common format which is used by the NCBI dbSNP database and is also widely used and easily parsed from different literature-sources. Publications out of the different databases and the genotypes of the users are associated with single SNPs by the Rs-ID. Allele and genotype frequencies are are updated regularly, based on the data present in openSNP.

@@

A flowchart of all services incorporated in openSNP and how users can upload or access the data is given in figure n+3. The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
